IML Term project paper

### 2022-12-22 paper.rmd
.

Our dataset includes 464 observations of this variable as well as corresponding observations to a collection of 103 other variables. These variables include 50 numeric variables which describe the mean measurement related to phenomenona such as carbon dioxide concentration, solar radiation and air temperature during a measurement day, 50 numeric variables which describe the corresponding standard deviations to these measurements for those same measurement days, an id listing which runs from 1 to 464, the date of each measurement as well as a logical variable describing whether the other measurements are partly bad. We also notice that some of the 50 types of measurements for which means and standard deviations are calculated are in fact related to the same phenomena but are simply measured at different points. For example, one of the measurements is the amount of carbon dioxide but this is measured at four different heights.

@@

Let's also calculate basic summary statistics for the other measurements

```{r}
summary(npf)
```

@@

## Generative models

@@

### Linear Discriminant Analysis

@@

### Quadratic Disriminant Analysis

@@

### Naive Bayes

@@

There are a few things of note about this model. First, it does not have an intercept, which implies that if all observations are at zero, the probability of an event would be 0.5. The validity of this assumption should be studied further in relation to the relevant qualities of NPF events. Second, we notice that we have especially high absolute values of coefficients for the variables related to the variables related to nitrous oxide, though said coefficients also have high standard errots. On top of this, we notice that mostly variables describing means have negative coefficients whereas variables describing standard deviations have positive coefficients. One possible interpretation could be that though lower values of the measurements in question throughout the day increase event probability, simply measuring the mean might not be sufficient, and instead questions of the variations throughout the day are relevant as well, specifically the positive coefficients imply that higher variation or for example higher extreme values within the day increase event probability.

@@

Second, in terms of our variable selection process through VIF and Wald tests, said process is rather simple and could be improved. The VIF iterations also remove corresponding mean and standard deviation variables even if their VIFs are smaller. This was done as the goal of the VIF process was to reduce the amount of phenomena to be observed to as small a collection as seems reasonable, and thus the notion of removing only one of the two variables related to the phenomenon felt slightly odd. Then again, this process was not followed with the Wald tests, as the used interpretation of the Wald tests is specifically more about coefficient values than about the optimization issues caused by multicolinearity, which we chose to discuss the VIF results as. The validity of this approach especially in terms of the VIF process should be questioned, even though results where we run VIF iteratively and only remove the variable that has the highest VIF value gives us a fairly similar model (overall 11 coefficients and an intercept with mostly the same set of explanatory variables).

Also, other variable selection processes such as Lasso (with less computationally expensive cross-validation) and PCA could be used equivalently. However, for the current report, we chose to use our current approach as our chosen method of assessing performance was LOO-CV and thus Lasso was computationally expensive for the multinomial model. Then again, a lower amount of folds in cross-validation could lead to better assessments of the accuracies, and thus be preferrable. We could have also used PCA but by using PCA, we'd lose the interpretability of the coefficients and when running the LOO-cv comparisons, we didn't see notable benefits in terms of accuracy or perplexity from using PCA on a larger collection of variables compared to our current approach of selecting a sufficiently small amount of explanatory variables.

>>>>

Our dataset includes 464 observations of this variable as well as corresponding observations to a collection of 103 other variables. These variables include 50 numeric variables which describe the mean measurement related to phenomena such as carbon dioxide concentration, solar radiation and air temperature during a measurement day, 50 numeric variables which describe the corresponding standard deviations to these measurements for those same measurement days, an id listing which runs from 1 to 464, the date of each measurement as well as a logical variable describing whether the other measurements are partly bad. We also notice that some of the 50 types of measurements for which means and standard deviations are calculated are in fact related to the same phenomena but are simply measured at different points. For example, one of the measurements is the amount of carbon dioxide but this is measured at four different heights.

@@

Let's also calculate basic summary statistics for the first few variables. The full summary of the whole dataset is available in appendix A.

```{r}
summary(npf[,1:4])
```

@@

### Generative models

@@

#### Linear Discriminant Analysis

@@

#### Quadratic Disriminant Analysis

@@

#### Naive Bayes

@@

There are a few things of note about this model. First, it does not have an intercept, which implies that if all observations are at zero, the probability of an event would be 0.5. The validity of this assumption should be studied further in relation to the relevant qualities of NPF events. Second, we notice that we have especially high absolute values of coefficients for the variables related to the variables related to nitrous oxide, though said coefficients also have high standard errots. On top of this, we notice that mostly variables describing means have positive coefficients whereas variables describing standard deviations have negative coefficients. One possible interpretation could be that though higher values of the measurements in question throughout the day increase event probability, simply measuring the mean might not be sufficient, and instead questions of the variations throughout the day are relevant as well, specifically the negative coefficient imply that smaller variation throughout the day increases event probability.

@@

Second, in terms of our variable selection process through VIF and Wald tests, said process is rather simple and could be improved. The VIF iterations also remove corresponding mean and standard deviation variables even if their VIFs are smaller. This was done as the goal of the VIF process was to reduce the amount of phenomena to be observed to as small a collection as seems reasonable, and thus the notion of removing only one of the two variables related to the phenomenon felt slightly odd. Then again, this process was not followed with the Wald tests, as the used interpretation of the Wald tests is specifically more about coefficient values than about the optimization issues caused by multicolinearity, which we chose to discuss the VIF results as. The validity of this approach especially in terms of the VIF process should be questioned.

Also, other variable selection processes such as Lasso (with less computationally expensive cross-validation) and PCA could be used equivalently. However, for the current report, we chose to use our current approach as we wanted both to preserve the interpretability of the coefficients that was missing from the PCA, and as our chosen method of assessing performance was LOO-CV, Lasso was computationally expensive for the multinomial model. Then again, a lower amount of folds in cross-validation could lead to better assessments of the accuracies, and thus be preferrable.
